*Prochlorococcus marinus* responses to light and oxygen

### 2024-06-06 Code/Manuscript_O2.Rmd

*Prochlorococcus marinus*, the smallest picocyanobacterium, comprises
multiple clades with distinct niches, currently across tropical and sub-tropical
oligotrophic ocean regions, including Oxygen Minimum Zones. Ocean
warming may open permissive temperatures in new, poleward photic
regimes, along with expanded Oxygen Minimum Zones. We used ocean metaproteomic
data to guide testing of *Prochlorococcus marinus* growth across a matrix of peak irradiances,
photoperiods, spectral bands and dissolved oxygen. MED4 from Clade HLI
requires greater than 4 h photoperiod, grows at 25 µmol O~2~ L^-1^ and
above, and exploits high cumulative diel photon doses. MED4, however, relies upon an alternative oxidase to
balance electron transport, which may exclude it from growth under our
lowest, 2.5 µmol O~2~ L^-1^, condition. SS120 from clade LLII/III is
restricted to low light under full 250 µmol O~2~ L^-1^, shows expanded
light exploitation under 25 µmol O~2~ L^-1^, but is excluded from growth
under 2.5 µmol O~2~ L^-1^. Intermediate oxygen suppresses the cost of
PSII photoinactivation, and possibly the enzymatic production of H~2~O~2~ in SS120,
which has limitations on genomic capacity for PSII and DNA repair. MIT9313 from
Clade LLIV is restricted to low blue irradiance under 250 µmol O~2~
L^-1^, but exploits much higher irradiance under red light, or under
lower O~2~ concentrations, conditions which slow photoinactivation of
PSII and production of reactive oxygen species. In warming oceans, range expansions and competition among clades will be governed not simply by light levels. Short photoperiods governed by latitude, temperate winter, and depth attenuation of light, will exclude clade HLI (including MED4). In contrast, clade LLII/III (including SS120), and especially clade LLIV (including MIT9313), may actually exploit higher light niches under expanding OMZ conditions, where low O~2~ relieves the stresses of oxidation stress and PSII photoinhibition.

@@

Conversely, *P. marinus* SS120 and MIT9313 lack genes encoding deoxyribodipyrimidine photolyase and DNA repair ligase (ATP), which may explain, in part, why these two strains cannot tolerate growth under full  250 µM [O~2~] and high light, as found at the ocean surface.  We show that *P. marinus* SS120 and MIT9313 do, in contrast, tolerate light levels representative of the near surface ocean at 25 µM [O~2~]. Our growth rate experiments did not include wavebands in the UV range, so we do not know whether the absence of deoxyribodipyrimidine photolyase would inhibit the growth rates of *P. marinus* SS120 and MIT9313  under low [O~2~] if exposed to full spectrum, ocean surface light. The protective effects of lower [O~2~], allowing these strains to grow at higher light, likely relates in part to suppression of DNA damage when generation of Reactive Oxygen Species (ROS) is suppressed at lower [O~2~]. NATL2A, a clade LLI, has been found near the ocean surface during deep ocean mixing [@malmstromTemporalDynamicsProchlorococcus2010] events. Malmstrom *et al*. [@malmstromTemporalDynamicsProchlorococcus2010] indeed attribute NATL2A tolerance to short exposures of high light to the presence of the genes encoding photolyase, also found in HL clades. The presence of deoxyribodipyrimidine photolyase but absence of DNA repair ligase (ATP)
supports why NATL2A tolerates limited exposure to high light and why NATL2A is unable to fully repair damaged DNA. *Prochlorococcus marinus* are highly susceptible to hydrogen peroxide (H~2~O~2~) toxicity as they lack genes which scavenge H~2~O~2~ molecules
[@morrisFacilitationRobustGrowth2008]. The small cell size of *P. marinus* allow the ROS, H~2~O~2~, to cross the cell membrane [@omarDiffusionalInteractionsMarine2022]; however, accumulation of extracellular H~2~O~2~ remains toxic to *P. marinus* [@morrisFacilitationRobustGrowth2008; @morrisDependenceCyanobacteriumProchlorococcus2011].

@@

XXXXDoug
The potential for niche expansion into temperate regions by *P. marinus*
varies depending on the season, which influences achieved underwater
photoperiods and light levels. Temperate summer delivers ~11 hours of
blue waveband light underwater, above the photic threshold of 20 µmol
photons m^-2^ s^-1^, while temperate spring/fall delivers 8 hours of
blue waveband light underwater, photoperiod ranges which are permissive
for growth of all three *P. marinus*. In contrast temperate winter
delivers only about 2 h of blue waveband light underwater above the
photic threshold, which precludes growth of MED4 and SS120, even if
winter waters reached permissive temperatures. MIT9313 and SS120 will be
excluded from near-surface growth niches by high PAR, unless OMZ zones
extend to the near surface (Table \@ref(tab:NichesTable).

@@

NichesMap_cap <- glue(" Potential Future Niches for *Prochlorococcus* clades in a warming ocean. 
The map shows current latitudinal distribution limits for *Prochlorococcus* (between green dashed lines), tropical and temperate  latitudinal regions (yellow bands). 
Light levels, photoperiods and spectral bands, potentially permissive for growth of *Prochlorococcus* clades are determined by the interactions of latitude, season and depth attenuation of light.
Potential clade occupancies under Full Oxygen (250 µM O~2~) are shown for 10-50 m and 100 m depths, in temperate vs. tropical regions, under depth-resolved photoperiods, approximating Winter Solstice, Equinox and Summer Solstice. Clade HLI is excluded from temperate depths by short photoperiods at Winter Solstice. Clades LLII/III are excluded from shallow depths by excessive light.
Low Oxygen Zones (25 µM O~2~) drastically alter these potential occupancy patterns by excluding Clade HLI from the temperate winter solstice at all depths, and by expanding the exploitation of higher light by Clade LLIV.
Potential clade occupancies for the southern temperate zone mirror the northern temperate zone occupancies.",
                       .sep = " ")

@@

Diverse *P. marinus* strains @billerProchlorococcusStructureFunction2015
differentially exploit potential photoregimes, both at the surface and
deep in the water column. Some *P. marinus* strains grow under low
oxygen environments, similar to OMZ. The LL clades we tested can
function as 'HL' in oxygen environments of 25 µM, and as low as 2.5 µM,
in the case of MIT9313.

@@

We analyzed growth rates for *P. marinus* clade HLI found near the ocean
surface; clade LLII/III found deep in the water column; and clade LLIV
also found in deep oceans, including OMZ, under a matrix of spectral
wavebands, irradiances, photoperiods and oxygen concentrations
approximating present day and hypothetical future niches.

*Prochlorococcus marinus* MED4 requires more than 4 h of light per day;
thus this strain will not exploit habitats typical of temperate winter
or light attenuated depths, even if water temperatures warm into the
clade HLI tolerance range. MED4 is also excluded from the lowest oxygen
habitats of 2.5 µM O~2~, but can, grow under OMZ regions with 25 µM
O~2~. Genomic (Fig \@ref(fig:ProchlorococcusEnzymeKms)) and
transcriptional analyses [@bergResponsesPsbAHli2011] suggest MED4 is
excluded from growth below \~ 25 µM O~2~ because it relies upon a
ubiquinol oxidase, non-electrogenic, to maintain oxidation/reduction
balance in the intersystem electron transport chain, with a K~m~ for
[O~2~] of \~25 µM O~2~. On the other hand, MED4 shows inducible
expression of FtsH isoforms [@bonisteelStrainSpecificDifferences2018],
to counter photoinactivation of PSII under higher PAR and [O~2~]
environments. However, photoinactivation imposes an increased cost of
growth upon MED4, since growth under red light, to lower
photoinactivation of PSII [@murphyPhotoinactivationPhotosystemII2017],
allows MED4 to achieve faster growth per absorbed photon than growth
under blue light. TARA Oceans Project data
[@delmontLinkingPangenomesMetagenomes2018] indeed reported presence of
*P. marinus* MED4-like genomes at depths ranging from 5 m to 90 m,
representing high to low blue light levels, in the Pacific South East
Ocean. Delmont and Eren [@delmontLinkingPangenomesMetagenomes2018] did
not analyze data from depths beyond the subsurface chlorophyll maximum
layer, nor did they report[O~2~] at depth. Our growth findings are
consistent with Fig \@ref(fig:CladeProchloroPhotosynthDepthO2)
showing PSII proteins annotated as MED4, clade HLI, at depths up to 200
meters, with O~2~ of \~15 µM.

@@

*Prochlorococcus marinus* MIT9313, a clade LLIV representative, shows
potential to inhabit future warmer temperate zones year-round, as it
grows under a 4 h photoperiod, expected in winter, or at
light-attenuated depths. MIT9313 demonstrates an unexpected tolerance to
higher light levels and cumulative diel PUR, but only under low oxygen
conditions of 25 µM and 2.5 µM (Fig \@ref(fig:MIT9313PARGAM)A), enabling
MIT9313 to grow in OMZ, even at depths closer to the surface. MIT9313
carries a gene encoding (S)-2-hydroxy-acid oxidase
[@cunaneCrystalStructureAnalysis2005], with a K~m~ for [O~2~] of \~ 250
µM (Fig \@ref(fig:ProchlorococcusEnzymeKms)), which produces
H~2~O~2~. Growth at lower [O~2~] may protect MIT9313 from
auto-intoxication from production of H~2~O~2~. We hypothesize that under
250 µM O~2~ and higher blue light, *P. marinus* MIT9313 suffers
photoinhibition, resulting in part from the inactivation of PSII caused
by the production of H~2~O~2~. This photoinhibition is compounded by
limited inducible repair for PSII, due to the absence of FtsH 1 and 2
expression in *P. marinus* MIT9313
[@bonisteelStrainSpecificDifferences2018]. MIT9313 shows remarkable
ability to thrive under very low [O~2~], potentially allowing it to
expand into broader ecological niches. Our results are supported by
Fig \@ref(fig:CladeProchloroPhotosynthDepthO2) showing PSI protein
subunits annotated as derived from MIT9313 detected at depths \> 120 m,
along with PSII subunits at depths from 50 m to 200 m in regions where
O~2~ was 15 µM. Bagby and Chisholm
[@bagbyResponseProchlorococcusVarying2015] suggest that O~2~ has a
protective role in *P. marinus* under lower carbon dioxide
environments when carbon fixation is limited. The deep water
environments typical for MIT9313 are relatively nutrient rich, and
*P. marinus* take up and metabolize various sugars
[@gomez-baenaGlucoseUptakeIts2008;
@munoz-marinGlucoseUptakeProchlorococcus2017;
@munoz-marinProchlorococcusCanUse2013] and amino acids
[@zubkovDepthRelatedAmino2004]. In future work we aim to test whether
MIT9313 is using photosynthesis to drive CO~2~ fixation in low O~2~
environments, or whether PSII generation of O~2~ acts as an electron
sink for respiration, using ATP for maintenance and to take up nutrients
from the surroundings. Partensky *et al*.
[@partenskyComparisonPhotosyntheticPerformances2018] indeed found that
in the low-light conditions found in the OMZ, MED4, SS120 and MIT9313
all became net O~2~ consumers, suggesting that low light levels
cause the respiratory chain to consume more O~2~ than the photosynthetic
electron transport chain generates, thus contributing to maintenance of
the low O~2~ environment.

In warming oceans *P. marinus* clades will differentially expand into new regions depending upon their different, interacting capacities to tolerate and exploit photoperiods, light levels, and [O~2~]. Competition among clades will be driven not simply by light levels. Clade HLI (including MED4) is excluded from short photoperiod regimes, typical of temperate winter, or light attenuated depths. In contrast, clade LLIV (including MIT9313) might actually exploit higher light niches under OMZ conditions, where low O~2~ relieves the stresses of oxidation stress and PSII photoinhibition.

>>>>

*Prochlorococcus marinus*, the smallest picocyanobacterium, comprises
multiple clades occupying distinct niches, currently across tropical and sub-tropical
oligotrophic ocean regions, including Oxygen Minimum Zones. Ocean
warming may open growth-permissive temperatures in new, poleward photic
regimes, along with expanded Oxygen Minimum Zones. We used ocean metaproteomic
data on current *Prochlorococcus marinus* niches, to guide testing of *Prochlorococcus marinus* growth across a matrix of peak irradiances,
photoperiods, spectral bands and dissolved oxygen. MED4 from Clade HLI
requires greater than 4 h photoperiod, grows at 25 µmol O~2~ L^-1^ and
above, and exploits high cumulative diel photon doses. MED4, however, relies upon an alternative oxidase to balance electron transport, which may exclude it from growth under our
lowest, 2.5 µmol O~2~ L^-1^, condition. SS120 from clade LLII/III is
restricted to low light under full 250 µmol O~2~ L^-1^, shows expanded
light exploitation under 25 µmol O~2~ L^-1^, but is excluded from growth
under 2.5 µmol O~2~ L^-1^. Intermediate oxygen suppresses the cost of
PSII photoinactivation, and possibly the enzymatic production of H~2~O~2~ in SS120,
which has limitations on genomic capacity for PSII and DNA repair. MIT9313 from
Clade LLIV is restricted to low blue irradiance under 250 µmol O~2~
L^-1^, but exploits much higher irradiance under red light, or under
lower O~2~ concentrations, conditions which slow photoinactivation of
PSII and production of reactive oxygen species. In warming oceans, range expansions and competition among clades will be governed not only by light levels. Short photoperiods governed by latitude, temperate winter, and depth attenuation of light, will exclude clade HLI (including MED4) from some habitats. In contrast, clade LLII/III (including SS120), and particularly clade LLIV (including MIT9313), may exploit higher light niches nearer the surface, under expanding OMZ conditions, where low O~2~ relieves the stresses of oxidation stress and PSII photoinhibition.

@@

Conversely, *P. marinus* SS120 and MIT9313 lack genes encoding deoxyribodipyrimidine photolyase and DNA repair ligase (ATP), which may explain, in part, why these two strains cannot tolerate growth under full  250 µM [O~2~] and high light, as found at the ocean surface.  *P. marinus* SS120 and MIT9313 do, in contrast, tolerate light levels representative of the near surface ocean at 25 µM [O~2~]. Our growth rate experiments did not include wavebands in the UV range, so we do not know whether the absence of deoxyribodipyrimidine photolyase would inhibit the growth rates of *P. marinus* SS120 and MIT9313  under low [O~2~] if exposed to full spectrum, ocean surface light. The protective effects of lower [O~2~], allowing these strains to grow at higher light, likely relates in part to suppression of DNA damage when generation of Reactive Oxygen Species (ROS) is suppressed at lower [O~2~]. NATL2A, a clade LLI, has been found near the ocean surface during deep ocean mixing [@malmstromTemporalDynamicsProchlorococcus2010] events. Malmstrom *et al*. [@malmstromTemporalDynamicsProchlorococcus2010] indeed attribute NATL2A tolerance of short exposures of high light to presence of genes encoding photolyase, also found in HL clades. The presence of deoxyribodipyrimidine photolyase but absence of DNA repair ligase (ATP)
supports why NATL2A tolerates limited exposure to high light, but is unable to fully repair damaged DNA. *Prochlorococcus marinus* are highly susceptible to hydrogen peroxide (H~2~O~2~) toxicity as they lack genes which scavenge H~2~O~2~ molecules [@morrisFacilitationRobustGrowth2008]. The small cell size of *P. marinus* allow the ROS, H~2~O~2~, to cross the cell membrane [@omarDiffusionalInteractionsMarine2022]; however, accumulation of extracellular H~2~O~2~ remains toxic to *P. marinus* [@morrisFacilitationRobustGrowth2008; @morrisDependenceCyanobacteriumProchlorococcus2011].

@@

Diverse *P. marinus* strains @billerProchlorococcusStructureFunction2015
differentially exploit potential photoregimes, both at the surface and
deeper in the water column. Some *P. marinus* strains grow under low
oxygen environments, similar to OMZ. The LLII and LLIV clade representatives we tested can
function as 'HL' in oxygen environments of 25 µM, and as low as 2.5 µM,
in the case of MIT9313.

Potential niche expansions of clades of *P. marinus* into temperate regions
vary depending on the season and the depth, which interact to govern both peak PAR levels and underwater photoperiods (Figure \@ref(fig:NichesMap). Temperate summer delivers ~12-14 hours of light, depending upon depth, above a photic threshold of 20 µmol photons m^-2^ s^-1^, while temperate spring/fall delivers ~9-11 hours; photoperiod ranges permissive for growth of all three *P. marinus*. In contrast temperate winter delivers shorter photoperiods which may exclude clade HLI.  In parallel, excess peak PAR will exclude clades LLII/III and LLIV from most shallow niches, even under permissive temperatures. Expansion of OMZ drastically alters these patterns, by allowing clade LLIV to exploit a much wider range of lights, extending towards near-surface niches in both tropical and temperate zones, under permissive temperatures.

@@

NichesMap_cap <- glue(" Potential Future Niches for *Prochlorococcus* clades in a warming ocean. 
The map shows current latitudinal distribution limits for *Prochlorococcus* (between green dashed lines), tropical and temperate  latitudinal regions (yellow bands). 
Light levels, photoperiods and spectral bands, potentially permissive for growth of *Prochlorococcus* clades are determined by the interactions of latitude, season and depth attenuation of light.
Potential clade occupancies under Full Oxygen (250 µM O~2~) are shown for 10-50 m and 100 m depths, in temperate vs. tropical regions, under depth-resolved photoperiods, approximating Winter Solstice, Equinox and Summer Solstice. Clade HLI is excluded from temperate depths by short photoperiods at Winter Solstice. Clades LLII/III are excluded from shallow depths by excessive light.
Low Oxygen Zones (25 µM O~2~) drastically alter these potential occupancy patterns by excluding clade HLI from the temperate winter solstice at all depths, and by expanding the exploitation of higher light by Clade LLIV.
Potential clade occupancies for the southern temperate zone mirror the northern temperate zone occupancies.",
                       .sep = " ")

@@

@@

We analyzed growth rates for *P. marinus* clade HLI usually found near the ocean
surface; clade LLII/III found deeper in the water column; and clade LLIV
also found in deep oceans, including current OMZ, under a matrix of spectral
wavebands, irradiances, photoperiods and oxygen concentrations
approximating present day and hypothetical future niches.

*Prochlorococcus marinus* MED4 requires more than 4 h of light per day;
thus this strain will not exploit habitats typical of temperate winter
or light attenuated depths, even if water temperatures warm into the
clade HLI tolerance range. MED4 is also excluded from the lowest oxygen
habitats of 2.5 µM O~2~, but can, grow under OMZ regions with 25 µM
O~2~. Genomic (Fig \@ref(fig:ProchlorococcusEnzymeKms)) and
transcriptional analyses [@bergResponsesPsbAHli2011] suggest MED4 is
excluded from growth below \~ 25 µM O~2~ because it relies upon a
ubiquinol oxidase, non-electrogenic, to maintain oxidation/reduction
balance in the intersystem electron transport chain, with a K~m~ for
[O~2~] of \~25 µM O~2~. On the other hand, MED4 shows inducible
expression of FtsH isoforms [@bonisteelStrainSpecificDifferences2018],
to counter photoinactivation of PSII under higher PAR and [O~2~]
environments. However, photoinactivation imposes an increased cost of
growth upon MED4, since growth under red light, to lower
photoinactivation of PSII [@murphyPhotoinactivationPhotosystemII2017],
allows MED4 to achieve faster growth per absorbed photon than growth
under blue light. TARA Oceans Project data
[@delmontLinkingPangenomesMetagenomes2018] indeed reported presence of
*P. marinus* MED4-like genomes at depths ranging from 5 m to 90 m,
representing high to low blue light levels, in the Pacific South East
Ocean. Our growth findings are consistent with Fig \@ref(fig:CladeProchloroPhotosynthDepthO2)
showing PSII proteins annotated as MED4, clade HLI, at depths up to 200
meters, with O~2~ of \~15 µM.

@@

*Prochlorococcus marinus* MIT9313, a clade LLIV representative, shows
potential to inhabit future warmer temperate zones year-round, as it
grows even under a 4 h photoperiod, expected in winter, or at
light-attenuated depths. MIT9313 demonstrates an unexpected tolerance to
higher light levels and cumulative diel PUR, but only under low oxygen
conditions of 25 µM and 2.5 µM (Fig \@ref(fig:MIT9313PARGAM)A), enabling
MIT9313 to grow in OMZ, even at depths closer to the surface. MIT9313
carries a gene encoding (S)-2-hydroxy-acid oxidase
[@cunaneCrystalStructureAnalysis2005], with a K~m~ for [O~2~] of \~ 250
µM (Fig \@ref(fig:ProchlorococcusEnzymeKms)), which produces
H~2~O~2~. Growth at lower [O~2~] may protect MIT9313 from
auto-intoxication from production of H~2~O~2~. We hypothesize that under
250 µM O~2~ and higher blue light, *P. marinus* MIT9313 suffers
photoinhibition, resulting in part from the inactivation of PSII caused
by the production of H~2~O~2~. This photoinhibition is compounded by
limited inducible repair for PSII, due to the absence of FtsH 1 and 2
expression in *P. marinus* MIT9313
[@bonisteelStrainSpecificDifferences2018]. MIT9313 shows remarkable
ability to thrive under very low [O~2~], potentially allowing it to
expand into broader ecological niches. Our results are supported by
Fig \@ref(fig:CladeProchloroPhotosynthDepthO2) showing protein
subunits derived from MIT9313 detected frequently at depths \> 120 m in regions where
O~2~ was 15 µM. Bagby and Chisholm
[@bagbyResponseProchlorococcusVarying2015] suggest that O~2~ has a
protective role in *P. marinus* under lower carbon dioxide
environments when carbon fixation is limited. The deep water
environments typical for MIT9313 are relatively nutrient rich, and
*P. marinus* take up and metabolize various sugars
[@gomez-baenaGlucoseUptakeIts2008;
@munoz-marinGlucoseUptakeProchlorococcus2017;
@munoz-marinProchlorococcusCanUse2013] and amino acids
[@zubkovDepthRelatedAmino2004]. In future work could test whether
MIT9313 is using photosynthesis to drive CO~2~ fixation in low O~2~
environments, or whether PSII generation of O~2~ acts as an electron
sink for respiration, using ATP for maintenance and to take up nutrients
from the surroundings. Partensky *et al*.
[@partenskyComparisonPhotosyntheticPerformances2018] indeed found that
in the low-light conditions found in the OMZ, MED4, SS120 and MIT9313
all became net O~2~ consumers, suggesting that low light levels
cause the respiratory chain to consume more O~2~ than the photosynthetic
electron transport chain generates, thus contributing to maintenance of
the low O~2~ environment.

In warming oceans *P. marinus* clades will differentially expand into new regions depending upon their different, interacting capacities to tolerate and exploit photoperiods, light levels, and [O~2~]. Competition among clades will be driven not simply by light levels. Clade HLI (including MED4) is excluded from short photoperiod regimes, typical of temperate winter, or light attenuated depths. In contrast, clade LLIV (including MIT9313) may exploit higher light niches under OMZ conditions, where low O~2~ relieves the stresses of oxidation stress and PSII photoinhibition.
